Chemical tagging can work: \\ Identification of stellar phase-space structures \\ purely by chemical-abundance similarity

### 2016-01-19 papers/clusters.tex
figure caption

\begin{abstract}
% Context:
Chemical tagging promises to robustly group co-natal, spatially extended stars by their present-day  photospheric chemical abundances.
This idea has not previously yielded scientific successes, probably because of the
noise and incompleteness in chemical-abundance measurements.
However, we have succeeded in substantially improving spectroscopic measurements with \thecannon: 
delivering $\geq$15 individual abundances for $\sim10^5$ stars with precisions around 0.04~dex.
% Aims:
We test the chemical-tagging hypothesis by looking at clusters in abundance space
and confirming that they are clustered in phase space.
% Methods:
We identify (by the k-means algorithm) overdensities of stars observed by the \apogee\ spectroscopic survey
in a 15-dimensional chemical-abundance space delivered by \thecannon,
and plot the associated stars in phase space.
We use \emph{only} abundance-space information (no positional information) to identify stellar groups.
% Results:
We find that clusters in abundance space are indeed clusters in phase space.
We recover some known clusters and find other interesting structures.
This confirms the chemical-tagging hypothesis and verifies the precision of abundances delivered by \thecannon.
This is the first-ever project to identify phase-space structures by blind search purely in abundance space;
the prospects for future data sets are very good.
\end{abstract}

@@

\begin{figure}[!bp]
\includegraphics[width=1.5\figwidth]{./figs/clusters_0256.png}
\caption{The distribution of membership and density for the 256
  abundance-space clusters returned by the k-means algorithm at
  $K=256$.  The densest cluster (which is displayed in more detail in 
  \figurename~\ref{fig:M13}) is far more dense than the second
  densest.\label{fig:densities}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0253}
\caption{Same as \figurename~\ref{fig:all} except that the entire
  sample has been made gray and the members of the most dense
  $K=256$ abundance-space cluster have been rendered as unique, prominent triangles.
  The lower-right plot shows this cluster (circle) in context of the other clusters (dots).
  This cluster, which was identified only in abundance space (six
  projections of which are the top six panels in this \figurename),
  turns out to be dominated by the halo globular cluster
  M13.\label{fig:M13}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0034}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by the disrupting globular
  cluster Palomar 5.\label{fig:Pal5}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0177}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by the Sagittarius star
  cloud.\label{fig:Sgr}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0010}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by a high-velocity-dispersion
  structure in the Galaxy halo.\label{fig:halo}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0141}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by a very thin young stellar
  structure in the Galaxy disk.\label{fig:disk}}
\end{figure}

@@

@@

The abundances of known clusters that we identify by k-means are in
excellent agreement with the literature. For the group shown in Figure \ref{fig:M13},
which we later attribute to M13, we find a mean of [Fe/H]$ = -1.55$ (DWH CHECK).
The mean and spread agree well with other works dedicated to studying this cluster
\citep{Kraft_1992,Cohen_Melendez_2005,Johnson_Pilachowski_2012}, and compiled catalogs of
globular cluster properties \citep{Harris}. The detailed abundances are also
consistent: we find a correlation in [C/Fe] and [N/Fe] abundances \citep{Smith_2005}, and their
projection in Figure \ref{fig:M13} show a spread that reflects deep mixing
along the red giant branch \citep{Briley_2004}. We find that the light element
abundances (most notably Mg--Al) for this cluster are anti-correlated, as
expected from other studies \citep[e.g.,][]{gratton}. However we find a
smaller spread in [O/Fe] than reported by others who have looked exclusively
at M13 \citep{Johnson_Pilachowski_2012}.  As we discuss above, this is expected:
k-means is by no means optimal for identifying arbitrary-shaped structures
in high-dimensional space, and the first stars that would be `discarded' from
this group would be those with the most extreme abundances: low [O/Fe] and
high [Na/Fe].

% Palomar 5
There are scant abundance measurements reported for Palomar 5, for which
most stars in the cluster in Figure \ref{fig:Pal5} belong. We find a mean
metallicity of [Fe/H]$ = -1.3$ (DWH CHECK), which lies between the only
photometric \citep{Geisler_1997} and spectroscopic \citep{Smith_1985,Smith_2002} abundance
measurements for this cluster.  The anti-correlation in C--N and Na--O abundance
have been seen elsewhere \citep[][, respectively]{Smith_1985,Smith_2002}, although
to our knowledge the Mg--Al anti-correlation has never been shown in this cluster.
Indeed, this work contains the largest number of abundances for Palomar 5,
permitting a thorough search of tidally disrupted members \citep[e.g.,][]{Kuzma_2015} 
that have retained the unique cluster abundance signature .

% ARC: Expand on why these particular abundance projections, with
%	comparisons to the literature.

@@

\clearpage
\begin{thebibliography}{24}\raggedright
\bibitem[Alam \etal(2015)]{dr12}
Alam,~S., Albareti,~F.~D., Allende~Prieto,~C., \etal, 2015, \apjs, 219, 12
\bibitem[Bishop(2006)]{bishop}
Bishop,~C.~M., 2006, \textit{Pattern Recognition and Machine Learning}, Springer, New York
\bibitem[Blanco-Cuaresma \etal(2015)]{blancocuaresma}
Blanco-Cuaresma,~S., Soubiran,~C., Heiter,~U., \etal, 2015, \aap, 577, A47
\bibitem[Bovy(2015)]{bovy}
Bovy,~J., 2015, arXiv:1510.06745
 \bibitem[Briley \etal(2004)]{Briley_2004} Briley, M.~M., Cohen, 
J.~G., \& Stetson, P.~B.\ 2004, \aj, 127, 1579 
\bibitem[Casey \etal(2016)]{casey16}
Casey,~A.~R., Hogg,~D.~W., Ness,~M., Rix,~H.-W., 2016, in preparation
\bibitem[Cohen \& Mel{\'e}ndez(2005)]{Cohen_Melendez_2005} Cohen, J.~G., \& Mel{\'e}ndez, J.\ 2005, \aj, 129, 303 
\bibitem[De~Silva \etal(2015)]{galah}
De~Silva,~G.~M., Freeman,~K.~C., Bland-Hawthorn,~J., \etal, 2015, \mnras, 449, 2604
\bibitem[Font \etal(2006)]{font}
Font,~A.~S., Johnston,~K.~V., Bullock,~J.~S., Robertson,~B.~E., 2006, \apj, 638, 585 
\bibitem[Freeman \& Bland-Hawthorn(2002)]{freeman}
Freeman,~K., Bland-Hawthorn,~J., 2002, \araa, 40, 487
\bibitem[Garc{\'{\i}}a~P{\'e}rez \etal(2015)]{aspcap}
Garc{\'{\i}}a~P{\'e}rez,~A.~E., Allende~Prieto,~C., Holtzman,~J.~A., \etal, 2015, arXiv:1510.07635
\bibitem[Geisler \etal(1997)]{Geisler_1997} Geisler, D., Claria, 
J.~J., \& Minniti, D.\ 1997, \pasp, 109, 799 
\bibitem[Gilmore \etal(2012)]{gaiaeso}
Gilmore,~G., Randich,~S., Asplund,~M., \etal, 2012, The Messenger, 147, 25 
\bibitem[Gratton \etal(2012)]{gratton}
Gratton,~R.~G., Carretta,~E., Bragaglia,~A., 2012, \aapr, 20, 50
\bibitem[Hunter(2007)]{matplotlib}
Hunter,~J.~D., 2007, Computing in Science \& Engineering, 9, 90
\bibitem[Jofr{\'e} \etal(2015)]{jofre}
Jofr{\'e},~P., M{\"a}dler,~T., Gilmore,~G., \etal, 2015, \mnras, 453, 1428 
\bibitem[Johnson \& Pilachowski(2012)]{Johnson_Pilachowski_2012} Johnson, C.~I., \& Pilachowski, C.~A.\ 2012, \apjl, 754, L38 
\bibitem[Kraft \etal(1992)]{Kraft_1992} Kraft, R.~P., Sneden, C., Langer, G.~E., \& Prosser, C.~F.\ 1992, \aj, 104, 645 
\bibitem[Kuzma \etal(2015)]{Kuzma_2015} Kuzma, P.~B., Da Costa, 
G.~S., Keller, S.~C., \& Maunder, E.\ 2015, \mnras, 446, 3297 
\bibitem[Majewski \etal(2015)]{apogee}
Majewski,~S.~R., Schiavon,~R.~P., Frinchaboy,~P.~M., \etal, 2015, arXiv:1509.05420
\bibitem[Martell(2015)]{martell}
Martell,~S.~L., 2015, arXiv:1507.00079
\bibitem[Martig \etal(2016)]{martig}
Martig,~M., Fouesneau,~M., Rix,~H.-W., \etal, 2016, \mnras, 456, 3655
\bibitem[Melendez \etal(2014)]{melendez}
Mel{\'e}ndez,~J., Ram{\'{\i}}rez,~I., Karakas,~A.~I., \etal, 2014, \apj, 791, 14
\bibitem[M{\'e}sz{\'a}ros \etal(2015)]{meszaros}
M{\'e}sz{\'a}ros,~S., Martell,~S.~L., Shetrone,~M., \etal, 2015, \aj, 149, 153
\bibitem[Mitschang \etal(2014)]{mitschang}
Mitschang,~A.~W., De~Silva,~G., Zucker,~D.~B., \etal, 2014, \mnras, 438, 2753
\bibitem[Ness \etal(2015)]{thecannon}
Ness,~M., Hogg,~D.~W., Rix,~H.-W., Ho,~A.~Y.~Q., Zasowski,~G., 2015, \apj, 808, 16
\bibitem[Ness \etal(2016a)]{ages}
Ness,~M., Hogg,~D.~W., Rix,~H.-W., \etal, 2016a, \apj, in press (arXiv:1511.08204)
\bibitem[Ness \etal(2016b)]{ness16}
Ness,~M., Hogg,~D.~W., Casey,~A.~R., Rix,~H.-W., 2016b, in preparation
\bibitem[Pedregosa \etal(2011)]{sklearn}
Pedregosa,~F., Varoquaux,~G., Gramfort,~A., \etal, 2011, Journal of Machine Learning Research, 12, 2825
\bibitem[Quillen \etal(2015)]{quillen}
Quillen,~A.~C., Anguiano,~B., De~Silva,~G., \etal, 2015, \mnras, 450, 2354
\bibitem[Ro{\v s}kar \etal(2008)]{roskar}
Ro{\v s}kar,~R., Debattista,~V.~P., Quinn,~T.~R., Stinson,~G.~S., Wadsley,~J., 2008, \apjl, 684, L79
\bibitem[Smith(1985)]{Smith_1985} Smith, G.~H.\ 1985, \apj, 298, 
249 
\bibitem[Smith \etal(2002)]{Smith_2002} Smith, G.~H., Sneden, C., 
\& Kraft, R.~P.\ 2002, \aj, 123, 1502 
\bibitem[Smith \etal(2005)]{Smith_2005} Smith, G.~H., Briley, 
M.~M., \& Harbeck, D.\ 2005, \aj, 129, 1589
\bibitem[Ting \etal(2016)]{ting}
Ting,~Y.-S., Conroy,~C., Rix,~H.-W., 2016, \apj, in press (arXiv:1507.07563)
\bibitem[van~der~Walt \etal(2011)]{numpy}
van~der~Walt,~S., Colbert,~S.~C., Varoquaux,~G., 2011, Computing in Science \& Engineering, 13, 22
\bibitem[Wegg \etal(2015)]{wegg}
Wegg,~C., Gerhard,~O., Portail,~M., 2015, \mnras, 450, 4050
\end{thebibliography}
\end{document}

>>>>

\begin{abstract}
% Context:
The chemical-tagging hypothesis is that every star ought to carry---%
in its detailed chemical abundances---%
an informative signature of its molecular-cloud birth location and time.
This idea has not previously yielded scientific successes, probably because of the
noise and incompleteness in chemical-abundance measurements.
However, we have succeeded in substantially improving spectroscopic measurements with \thecannon\ 
(now delivering $\geq$15 individual abundances for $\sim10^5$ stars with precisions around 0.04~dex).
% Aims:
We test the chemical-tagging hypothesis by looking at clusters in abundance space
and confirming that they are clustered in phase space.
% Methods:
We identify (by the k-means algorithm) overdensities of stars observed by the \apogee\ spectroscopic survey
in a 15-dimensional chemical-abundance space delivered by \thecannon,
and plot the associated stars in phase space.
We use \emph{only} abundance-space information (no positional information) to identify stellar groups.
% Results:
We find that clusters in abundance space are indeed clusters in phase space.
We recover some known clusters and find other interesting structures.
This confirms the chemical-tagging hypothesis and verifies the precision of abundances delivered by \thecannon.
This is the first-ever project to identify phase-space structures by blind search purely in abundance space;
the prospects for future data sets are very good.
\end{abstract}

@@

\begin{figure}[!bp]
\includegraphics[width=1.5\figwidth]{./figs/clusters_0256.png}
\caption{The distribution of membership and density for the 256
  abundance-space clusters returned by the k-means algorithm at
  $K=256$.  The densest cluster (which is displayed in more detail in 
  \figurename~\ref{fig:M13}) is far more dense than the second
  densest.\label{fig:densities}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0253}
\caption{Same as \figurename~\ref{fig:all} except that the entire
  sample has been made gray and the members of the most dense
  $K=256$ abundance-space cluster have been rendered as unique, prominent triangles
  (color is velocity rank, orientation is $\logg$ rank).
  The lower-right plot shows this cluster (circle) in context of the other clusters (dots).
  This cluster, which was identified only in abundance space (six
  projections of which are the top six panels in this \figurename),
  turns out to be dominated by the halo globular cluster
  M13.\label{fig:M13}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0034}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by the disrupting globular
  cluster Palomar 5.\label{fig:Pal5}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0177}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by the Sagittarius star
  cloud.\label{fig:Sgr}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0010}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by a high-velocity-dispersion
  structure in the Galaxy halo.\label{fig:halo}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0141}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by a very thin young stellar
  structure in the Galaxy disk.\label{fig:disk}}
\end{figure}

@@

The abundances we find for clusters identified in k-space are in excellent
agreement with detailed studies that exist in the literature. The mean
metallicity for stars we identify in abundance space that are later attributable
to M13 have a mean of [Fe/H]$ = -1.55$, in agreement with other works \citep{Kraft_1992,Cohen_Melendez,Johnson_Pilachowski} and the \citet{Harris} catalog.
The detailed abundances are also in agreement: we find a correlation in 
[C/Fe] and [N/Fe] abundances, and an intrinsic spreads in each element
that reflects deep mixing along the red giant branch.

% M13
% ANDY IS EDITING AND COLLECTING CRAP HERE.
We have a smaller spread in [O/Fe] than Johnson & Pilachowski (http://iopscience.iop.org/article/10.1088/2041-8205/754/2/L38/pdf)
Theirs is -1 to +1

The mean of [Na/Fe] and [O/Fe] is consistent with J & P.

%\bibitem[Johnson 
%\& Pilachowski(2012)]{2012ApJ...754L..38J} Johnson, C.~I., \& Pilachowski, C.~A.\ 2012, \apjl, 754, L38

Our [N/Fe] and [C/Fe] overlaps perfectly with Smith et al (2005). Our scatter in [C/Fe] is slightly less: this is because those with the lowest [C/Fe] ratios are the brightest cluster members, which are not in this cluster (e.g., lowest Teff is ~4500) -- did APGOEE observe them?

%\bibitem[Smith et al.(2005)]{2005AJ....129.1589S} Smith, G.~H., Briley, 
%M.~M., \& Harbeck, D.\ 2005, \aj, 129, 1589

C-depletion wrt Teff (due to mixing) also seen in

 \bibitem[Briley et al.(2004)]{2004AJ....127.1579B} Briley, M.~M., Cohen, 
J.~G., \& Stetson, P.~B.\ 2004, \aj, 127, 1579

% Abundances (Mg, Al, Na, O, Fe, and probably other projections) perfectly consistent with 
%\bibitem[Cohen 
%\& Mel{\'e}ndez(2005)]{2005AJ....129..303C} Cohen, J.~G., \& Mel{\'e}ndez, J.\ 2005, \aj, 129, 303

and

%\bibitem[Kraft et al.(1992)]{1992AJ....104..645K} Kraft, R.~P., Sneden, C., 
%Langer, G.~E., \& Prosser, C.~F.\ 1992, \aj, 104, 645 
% except perhaps we have fewer low [O/Fe] objects than Kraft et al.

% Palomar 5

ARC: How do the abundances that we find in these known clusters compare
	with the literature?

% Overall abundances (mean [fE/h])
% individual abundances
% anti-correlations
% ARC: Expand on why these particular abundance projections, with
%	comparisons to the literature.

@@

@@

\clearpage
\begin{thebibliography}{24}\raggedright
\bibitem[Alam \etal(2015)]{dr12}
Alam,~S., Albareti,~F.~D., Allende~Prieto,~C., \etal, 2015, \apjs, 219, 12
\bibitem[Bishop(2006)]{bishop}
Bishop,~C.~M., 2006, \textit{Pattern Recognition and Machine Learning}, Springer, New York
\bibitem[Blanco-Cuaresma \etal(2015)]{blancocuaresma}
Blanco-Cuaresma,~S., Soubiran,~C., Heiter,~U., \etal, 2015, \aap, 577, A47
\bibitem[Bovy(2015)]{bovy}
Bovy,~J., 2015, arXiv:1510.06745
\bibitem[Casey \etal(2016)]{casey16}
Casey,~A.~R., Hogg,~D.~W., Ness,~M., Rix,~H.-W., 2016, in preparation
\bibitem[De~Silva \etal(2015)]{galah}
De~Silva,~G.~M., Freeman,~K.~C., Bland-Hawthorn,~J., \etal, 2015, \mnras, 449, 2604
\bibitem[Font \etal(2006)]{font}
Font,~A.~S., Johnston,~K.~V., Bullock,~J.~S., Robertson,~B.~E., 2006, \apj, 638, 585 
\bibitem[Freeman \& Bland-Hawthorn(2002)]{freeman}
Freeman,~K., Bland-Hawthorn,~J., 2002, \araa, 40, 487
\bibitem[Garc{\'{\i}}a~P{\'e}rez \etal(2015)]{aspcap}
Garc{\'{\i}}a~P{\'e}rez,~A.~E., Allende~Prieto,~C., Holtzman,~J.~A., \etal, 2015, arXiv:1510.07635
\bibitem[Gilmore \etal(2012)]{gaiaeso}
Gilmore,~G., Randich,~S., Asplund,~M., \etal, 2012, The Messenger, 147, 25 
\bibitem[Gratton \etal(2012)]{gratton}
Gratton,~R.~G., Carretta,~E., Bragaglia,~A., 2012, \aapr, 20, 50
\bibitem[Hunter(2007)]{matplotlib}
Hunter,~J.~D., 2007, Computing in Science \& Engineering, 9, 90
\bibitem[Jofr{\'e} \etal(2015)]{jofre}
Jofr{\'e},~P., M{\"a}dler,~T., Gilmore,~G., \etal, 2015, \mnras, 453, 1428 
\bibitem[Majewski \etal(2015)]{apogee}
Majewski,~S.~R., Schiavon,~R.~P., Frinchaboy,~P.~M., \etal, 2015, arXiv:1509.05420
\bibitem[Martell(2015)]{martell}
Martell,~S.~L., 2015, arXiv:1507.00079
\bibitem[Martig \etal(2016)]{martig}
Martig,~M., Fouesneau,~M., Rix,~H.-W., \etal, 2016, \mnras, 456, 3655
\bibitem[Melendez \etal(2014)]{melendez}
Mel{\'e}ndez,~J., Ram{\'{\i}}rez,~I., Karakas,~A.~I., \etal, 2014, \apj, 791, 14
\bibitem[M{\'e}sz{\'a}ros \etal(2015)]{meszaros}
M{\'e}sz{\'a}ros,~S., Martell,~S.~L., Shetrone,~M., \etal, 2015, \aj, 149, 153
\bibitem[Mitschang \etal(2014)]{mitschang}
Mitschang,~A.~W., De~Silva,~G., Zucker,~D.~B., \etal, 2014, \mnras, 438, 2753
\bibitem[Ness \etal(2015)]{thecannon}
Ness,~M., Hogg,~D.~W., Rix,~H.-W., Ho,~A.~Y.~Q., Zasowski,~G., 2015, \apj, 808, 16
\bibitem[Ness \etal(2016a)]{ages}
Ness,~M., Hogg,~D.~W., Rix,~H.-W., \etal, 2016a, \apj, in press (arXiv:1511.08204)
\bibitem[Ness \etal(2016b)]{ness16}
Ness,~M., Hogg,~D.~W., Casey,~A.~R., Rix,~H.-W., 2016b, in preparation
\bibitem[Pedregosa \etal(2011)]{sklearn}
Pedregosa,~F., Varoquaux,~G., Gramfort,~A., \etal, 2011, Journal of Machine Learning Research, 12, 2825
\bibitem[Quillen \etal(2015)]{quillen}
Quillen,~A.~C., Anguiano,~B., De~Silva,~G., \etal, 2015, \mnras, 450, 2354
\bibitem[Ro{\v s}kar \etal(2008)]{roskar}
Ro{\v s}kar,~R., Debattista,~V.~P., Quinn,~T.~R., Stinson,~G.~S., Wadsley,~J., 2008, \apjl, 684, L79
\bibitem[Ting \etal(2016)]{ting}
Ting,~Y.-S., Conroy,~C., Rix,~H.-W., 2016, \apj, in press (arXiv:1507.07563)
\bibitem[van~der~Walt \etal(2011)]{numpy}
van~der~Walt,~S., Colbert,~S.~C., Varoquaux,~G., 2011, Computing in Science \& Engineering, 13, 22
\bibitem[Wegg \etal(2015)]{wegg}
Wegg,~C., Gerhard,~O., Portail,~M., 2015, \mnras, 450, 4050
\end{thebibliography}
\end{document}
